Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-10-07 agda.tex
enhanced determinism section

% !TEX root = hazelnut-popl17.tex
So far, we have given an overview of the most important judgements and
rules in the semantics of Hazelnut, and stated the critical metatheorem,
Sensibility, and several auxiliary ``checks''. In a few cases, we have
informally sketched out why these metatheorems will hold. In order to
formally establish that our design meets our stated objectives, we have
mechanized the semantics of Hazelnut using the Agda proof
assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.)

This development is available in the supplemental material. It contains the
full set of rules given in the paper, and proofs of all of the theorems
described. It also contains the rules needed for the extension to the
language described in Sec. \ref{sec:extending}, and extends the proofs to
the constructs.

>>>>

% !TEX root = hazelnut-popl17.tex
% So far, we have given an overview of the most important judgements and
% rules in the semantics of Hazelnut, and stated the critical metatheorem,
% Sensibility, and several auxiliary ``checks''. In a few cases, we have
% informally sketched out why these metatheorems will hold.

In order to
formally establish that our design meets our stated objectives, we have
mechanized the semantics and metatheory of Hazelnut as described above (and more concisely in the appendix) using the Agda proof
assistant \cite{norell:thesis} (also see the Agda Wiki, hosted
at \url{http://wiki.portal.chalmers.se/agda/}.) This development is available in the supplemental material. The mechanization also includes the  extension to the
language described in Sec. \ref{sec:extending}.

### 2016-10-07 formally.tex
enhanced determinism section

\subsection{Determinism}
A last useful property to consider is \emph{action determinism}, i.e. that
performing an action produces a unique result. Formally, this would be
established as follows:
\begin{theorem}[Action Determinism]
  \label{thrm:actdet} All of the following hold:
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
    $\zexp' = \zexp''$ and $\htau' = \htau''$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
    \zexp''$.
  \end{enumerate}
\end{theorem}
This is not a theorem of the system exactly as described so far. The reason
is quite subtle: the semantics of two construction actions,
$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
analytic case than they do in the synthetic case -- the difference has to
do with what the generated ascription contains. Both cases can technically
``fire'' due to action subsumption, but one corresponds to an over-eager
use of subsumption: it's intended to be a rule of last resort.

The mechanization defines a predicate on action derivations that bans
exactly these over-eager uses of action subsumption, gives a mapping from
any derivation into one that matches the source and action and satisfies
the predicate, and proves determinism as stated above but only for
derivations modulo that predicate. This fix is analogous to the difference
between a chaining and focused logic in proof theory, which correspond
closely with bidirectional type checking systems, and therefore is to be
expected.\cite{Simmons11tr}

>>>>

\subsection{Determinism}
A last useful property to consider is \emph{action determinism}, i.e. that
performing an action produces a unique result. The following theorem establishes determinism for type actions:
\begin{theorem}[Type Action Determinism]
\label{thrm:type-actdet} If $\performTyp{\ztau}{\alpha}{\ztau'}$ and
    $\performTyp{\ztau}{\alpha}{\ztau''}$ then $\ztau'=\ztau''$.
\end{theorem}

The corresponding theorem for expression actions would be stated as follows:
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
    $\performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ then
    $\zexp' = \zexp''$ and $\htau' = \htau''$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
    $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ then $\zexp' =
    \zexp''$.
  \end{enumerate}

This is not a theorem of the system as described so far. The reason
is somewhat subtle: two construction actions,
$\aConstruct{\fasc}$ and $\aConstruct{\flam{x}}$, behave differently in the
analytic case than they do in the synthetic case. The problem is that both rules can  
``fire'' when considering a Z-expression in analytic position due to action subsumption. This is a technically valid but ``morally'' invalid use of action subsumption: subsumption is included in the system to be used as a rule of last resort, i.e. it should only be applied when no other analytic action rule can fire.

There are several possible ways to address this problem. One approach would be to modify the judgement forms to internalize this notion of ``rule of last resort''. This approach is related to focusing from proof theory \cite{Simmons11tr}. However, this approach would substantially complicate our presentation of the system.

The approach that we take leaves the system unchanged. Instead, we define predicates over \emph{derivations} of the expression action judgements that excludes those derivations that apply subsumption prematurely, i.e. when another rule could have been applied. We call such derivations \emph{subsumption-minimal derivations}. We can establish determinism for subsumption-minimal derivations.
\begin{theorem}[Expression Action Determinism] ~
  \begin{enumerate}[itemsep=0px,partopsep=0px,topsep=0px]
  \item If $\hsyn{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\mathcal{D}_1 : \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}{\htau'}$ and
    $\mathcal{D}_2 : \performSyn{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}{\htau''}$ and $\subminsyn{\mathcal{D}_1}$ and $\subminsyn{\mathcal{D}_2}$ then
    $\zexp' = \zexp''$ and $\htau' = \htau''$.
  \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
    $\mathcal{D}_1 : \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ and
    $\mathcal{D}_2 : \performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp''}$ and $\subminana{\mathcal{D}_1}$ and $\subminana{\mathcal{D}_2}$ then $\zexp' =
    \zexp''$.
  \end{enumerate}
\end{theorem}

The mechanization, discussed next, defines the predicates $\subminsyn{\mathcal{D}}$ and $\subminana{\mathcal{D}}$. In addition, it defines a mapping from
any derivation into a corresponding subsumption-minimal derivation. Implementors of Hazelnut are expected only to implement this subsumption-minimal fragment.
